---
title: "The Holy Experiment"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
properties <- read.csv('ReligiousProperties_coded.csv')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#adds a century completed column
properties$CENTURY.COMPLETED <- cut(properties$YEAR.ORIGINAL.BUILDING.COMPLETED, 
                                    c(1699, 1799, 1899, 1999))
properties$CENTURY.COMPLETED <- as.character(properties$CENTURY.COMPLETED)

properties$CENTURY.COMPLETED[properties$CENTURY.COMPLETED == 
                               '(1.8e+03,1.9e+03]'] <- '19C'
properties$CENTURY.COMPLETED[properties$CENTURY.COMPLETED == 
                               '(1.9e+03,2e+03]'] <- '20C'
properties$CENTURY.COMPLETED[properties$CENTURY.COMPLETED == 
                               '(1.7e+03,1.8e+03]'] <- '18C'

```

```{r message=FALSE, warning=FALSE, echo=FALSE}

#selects only stuff I'm potentially interested in

clean.properties <- subset(properties, select = c('CENTURY.COMPLETED', 'CURRENT.CONGREGATION', 'ZIP.CODE', 'YEAR.ORIGINAL.BUILDING.COMPLETED', 'Phila.Register..District.', 'National.Register..Ind.', 
                                              'National.Register..District.'))

```

##Original Completion Dates

```{r message=FALSE, warning=FALSE, echo=FALSE}
summary(clean.properties$YEAR.ORIGINAL.BUILDING.COMPLETED)

```

The earliest building in the table was completed in 1700. Philadelphia was 
originally settled in 1682, so this date is realistic for the area. The most
recent building completed was in 1981, which seems a bit late, but it may have
 architectural significance that would put it on the register.

```{r message=FALSE, warning=FALSE, echo=FALSE}
# Histogram of buildings completed across all years. 
ggplot(aes(x=YEAR.ORIGINAL.BUILDING.COMPLETED), data=clean.properties) +
  geom_histogram(binwidth=10) 
```

The highest completions were in the late 19th and early 20th centuries, which
 coincide with a boom in Philadelphia's outlying neighborhoods, largely due to
  the increased availability of public transit, and the formation of streetcar
   suburbs. 

```{r message=FALSE, warning=FALSE, echo=FALSE}

#Histogram of buildings completed broken down by zipcode
ggplot(aes(x=YEAR.ORIGINAL.BUILDING.COMPLETED), data=clean.properties) +
  geom_histogram(binwidth=10) +
  facet_wrap(~ZIP.CODE)
```

Completion years, faceted by zipcode. This will eventually be a GIS map as well,
 but the really active areas can be relatively easily picked out.

```{r message=FALSE, warning=FALSE, echo=FALSE}
#FIGURE OUT HOW TO HAVE EACH ZIP REPRESENTED BY A BAR
ggplot(aes(x=factor(ZIP.CODE)), 
       data=clean.properties) +
  geom_bar(aes(fill=CENTURY.COMPLETED))  +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Showing the prevalance of 20th-century construction -- although some zips 
noticeably see more growth earlier. These are probably the areas closest to 
Center City, where the bulk of Philadelphians lived until the early 20th 
century.

#Analysis

Although this is probably a dataset best shown off by the accompanying 
(eventually) GIS maps, it gives a good overview of when churches were being 
built in Philadelphia. Although founded in the late 1600's, it's clear that 
people didn't really begin to move to outlying neighborhoods -- requiring 
churches, synagogues, etc. -- until the late 19th and early 20th centuries. 

The zip code breakdowns can also serve as an interesting peek into the diversity
 of Philadelphia's religious communities; this dataset represents a huge number
  of faiths, and it could be possible to trace movement of ethnic and national 
  groups, among others, by how churches move, when they were built, and where
   they were built. I hypothesize that zipcodes with lots of construction, 
   particularly in the 20th century, aren't just showing a high population, but
    a very diverse one as well.


## Building Reuse

#Original Faiths Examined

```{r message=FALSE, warning=FALSE, echo=FALSE}
relproperties <- subset(properties, select = c('CENTURY.COMPLETED', 'ZIP.CODE', 'YEAR.ORIGINAL.BUILDING.COMPLETED','ORIGINAL.FAITH', 'CURRENT.FAITH'))

#Histogram showing counts of original faiths
ggplot(aes(x=ORIGINAL.FAITH), data=relproperties) +
  geom_histogram(stat='count') +
  coord_flip()
```

Churches broken down by original faith. Catholic, Presbytherian, Episcopal and
 Baptist all have the highest numbers of churches. Of note are the faiths that
 have disappeared or been folded into others over time, including German 
 Calvinist, Swedenborgian, Schwenkfelder and American Brethren (which has a 
 presence in Pennsylvania still).

```{r message=FALSE, warning=FALSE, echo=FALSE}
# scatterplot of faith, location, colored by century
ggplot(aes(x=YEAR.ORIGINAL.BUILDING.COMPLETED, y=ORIGINAL.FAITH), 
       data=relproperties) +
  geom_point(aes(color=CENTURY.COMPLETED))

```

This graph shows the growth of various faiths over time, perhaps representing 
the changing demographics of Philadelphia. For example, until the mid-19th 
century, there were more Quaker meetinghouses than Roman Catholic churches.

#Analysis

The large number of faiths represented may well be unique to Philadelphia, or 
at least Pennsylvania, and speaks to the original purpose of the colony -- to
 allow free practice of religion. It's interesting to note that the faiths with
  the most churches generally don't take off until well into the 19th century. 

I'm also really interested by the American Brethren churches, of which one was
built in the late 18th century, and then none until the 20th century. 

Philadelphia's immigrant history can also be traced to when churches start 
appearing, as with the Ukrainian Catholic church in the late 20th century. (It
 should be noted that these are specifically churches that appear on an historic
  register, and can therefore almost be considered sample of the total churches 
  in Philadelphia. Real demographic analysis is probably best conducted with a 
  larger, more random sample or the entire population.)

#Adaptive Reuse

```{r message=FALSE, warning=FALSE, echo=FALSE}
count(relproperties, CURRENT.FAITH == 'ADAPTIVE REUSE')

reused.relproperties = subset(relproperties, relproperties$CURRENT.FAITH == 
                                'ADAPTIVE REUSE')
```

33 properties are recorded as having 'adaptive re-use'; they are still in use, 
but are no longer used for religious activities. The other 714 properties in the
dataset are religous buildings, although possibly under a different faith.

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot(aes(x=YEAR.ORIGINAL.BUILDING.COMPLETED), data=reused.relproperties) +
  geom_histogram(bins=50)
```

As with the dataset as a whole, the majority of buildings adaptively reused were
 originally completed in the late 19th and early 20th centuries.

```{r}
a1 <- ggplot(aes(x=factor(reused.relproperties$ZIP.CODE)), data=reused.relproperties) +
  geom_histogram(stat='count')

a2 <- ggplot(aes(x=factor(relproperties$ZIP.CODE)), data=relproperties) + 
  geom_histogram(stat='count') +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))


grid.arrange(a1, a2, ncol=1)
```

Reused properties (top) compared to all properties, broken down by zip code. 

19128 is a particularly interesting outlier -- it has one of the highest counts 
of adaptive re-use, but doesn't particularly stand out for property counts in 
the full dataset. 19104 and 19147 at least are on the higher end of overall 
number of religious properties.

#Properties that changed denomination

```{r message=FALSE, warning=FALSE, echo=FALSE}
change.properties <- subset(properties, properties$CHANGE.IN.FAITH.DENOMINATION 
                            == 'YES')
change.properties <- subset(change.properties, select = c('CENTURY.COMPLETED', 
  'ZIP.CODE', 'YEAR.ORIGINAL.BUILDING.COMPLETED','ORIGINAL.FAITH', 
  'CURRENT.FAITH'))
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot(aes(x=factor(ORIGINAL.FAITH), y=factor(CURRENT.FAITH)), 
           data=change.properties) + 
         geom_point() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Original faith versus current faith.
